\textbf{Microbiota predict \emph{Clostridium difficile} severity in germ-free mice colonized with human feces}

### 2016-12-01 submission/Manuscript.tex

\emph{Clostridium difficile} causes diarrheal disease when it
successfully colonizes a dysbiotic gut microbial community. Current
mouse models to study \emph{C. difficile} infection (CDI) rely on
pre-treatment with antibiotics to disrupt the mouse microbiome prior to
inoculation. This model does not allow for analysis of human-associated
microbial community members that modulate \emph{C. difficile}
colonization and expansion. To study human-associated microbes in the
context of CDI, we inoculated germ-free C57BL/6 mice with one of 16
human fecal samples from diarrheal or healthy donors and challenged with
C. difficile 14 days later. Five unique donor-mice combinations resulted
in severe CDI while the remaining 11 only experienced mild disease. Both
healthy and diarrheal donors were susceptible to colonization and severe
symptoms of CDI. To determine if specific microbes were associated with
disease severity outcomes, we built a classification Random Forest
machine learning model based on relative abundance data of the
communities prior to infection. The model identified a number of
bacterial populations associated with the development of severe CDI,
including \emph{Bacilliales, Ruminococcaceae, Ruminococcus,
Staphylococcus, Streptococcus and Bacteriodetes}. Additionally, a
regression model accurately predicted colonization levels of \emph{C.
difficile} at one to ten days post-infection. This model explained 99\%
of the variance in the number of CFU isolated from mouse stool. Members
of \emph{Lachnospiraceae, Parabacteroides, Bacteroidales, Bacteroidetes,
Porphyromonadaceae} and unclassified \emph{Bacteria} families were
predictive of future \emph{C. difficile} colonization levels. Finally,
challenging these mice with different strains of \emph{C. difficile}
revealed that susceptible human-associated microbial communities were
prone to severe disease independent of strain type. Taken together these
results suggest that human-associated microbial communities can be
recapitulated in germ-free mice and used to characterize dynamics of
CDI. Because both healthy and diarrheal patients were susceptible to
severe CDI, machine-learning models are useful to identify bacterial
populations that allow colonization and contribute to the development of
\emph{C. difficile} associated disease in humans.

@@

The power of the gnotobiotic models to study CDI has been further
realized by first inoculating germ-free mice and piglets with human
stool microbes. In one study, germ-free piglets were acutely colonized
with human feces for one week and then treated with tigecycline. After
challenge with C. difficile none of the antibiotic-treated piglets
succumbed to infection, while some of the untreated human-colonized pigs
did (11). In another study, germ-free mice colonized with human feces
were bred over several generations to create a cohort of mice with
identical human-derived microbiomes (14). These mice were subsequently
treated with a five-antibiotic cocktail to induce dysbiosis and then
were successfully colonized by \emph{C. difficile} (14). While
informative, these studies were limited in their use of only one human
donor as input inoculum. In order to best understand the impact of
\emph{C. difficile} pathogenesis on human disease, we must have a
laboratory model that allows for study of a variety of human-derived
microbiomes.

Here we designed a laboratory model that allows for studying and
modeling microbial coumminity interactions with \emph{C. difficile}
infection in mice with human derived microbiomes. To test the impact of
individual human microbiomes on CDI, we colonized germ-free mice with 16
different human stool donors. We then characterized human-associated
microbiome response to C. difficile challenge. Additionally, the use of
machine-learning models allowed us to build a predictive model that
classified ``at-risk'' microbiomes prior to infection with C. difficile.
These findings show that human-associated microbiomes can be at risk for
CDI even in the absence of antibiotics and that study of mice colonized
with human feces provides a range of clinical outcomes.

\textbf{also want something in the intro about severity maybe, and
asymptomatic colonization. or in discussion?}

@@

\textbf{Germ-free mice inoculated with human feces as model for C.
difficile infection.} To generate mice with human-derived microbiomes,
we inoculated one cage of gnotobiotic C57/BL6 mice with one of 16
different human fecal donors. Five donors were patients that had
diarrhea that was not attributable to \emph{C. difficile} infection
while 11 donors were healthy at time of donation. Stool from a patient
that was colonized with virulent \emph{C. difficile} was used as a
positive control. After inoculation with human stool, mice were allowed
to equilibrate for 14 days. Prior to infection, stool samples were taken
from each mouse to establish baseline. Then, the \emph{C. difficile}
strain isolated from the positive control patient's sample (strain 430)
was used to infect each mouse with 100 spores. Mice were monitored for
weight loss and clinical signs of disease. Fecal samples were taken to
enumerate \emph{C. difficile} CFU and for microbiome analysis every day
for up to 10 days post-infection (Fig 1A). To ensure that the donors we
selected represented a diverse array of human microbiomes, we sequenced
the 16S rRNA genes from donor fecal inocula. Ordination of the distances
between donor communities showed that the donors each had distinctly
different communities, independent of whether the sample came from a
sick or healthy person (Fig 1B). Likewise, the starting microbial
communities of the mice on day 0 were characterizing by sequencing of
fecal pellets DNA prior to infection. Ordination of all of the mouse
communities on day 0 shows that mice were similar to each other within
each cage and donor, but distinct from other donors (Fig 1C). This
result confirmed that human-associated microbes were able to colonize
gnotobiotic mice and provide distinct initial communities to test
\emph{C. difficile} dynamics.

\textbf{need to put in a line to reference supplemental table with RA in
it}

\textbf{C. difficile infection in mice with human-derived microbiota
cause a range of outcomes.} \emph{C. difficile} colonization was
monitored by daily plating of stool pellets for \emph{C. difficile} CFU.
Nearly all of the mice were colonized to 10\textsuperscript{5} --
10\textsuperscript{7} CFU by one day post-infection and remained
colonized at that level until the end of the experiment (Fig 2A). As one
indicator of disease, mouse weights were taken each day post-infection
and weight-loss was monitored alongside clinical signs of disease. When
mice were judged to be too ill to continue they were humanely
euthanized. Overall, disease phenotypes fell into two classes. Mice that
became severely ill and lost 20\% or more of their starting body weight
within one to two days post-infection were classified as ``severe''
whereas mice that were colonized with \emph{C. difficile} but did not
show signs of disease or severe weight loss were considered to have
``mild'' disease (Fig 2A, 2B). Interestingly, \emph{C. difficile} was
able to cause severe disease in both mice that had been colonized with
healthy stool and those colonized with diarrheal stool, suggesting
susceptibility to CDI is dependent on the composition of the starting
microbiome and not associated with donor clinical metadata.

\textbf{Microbes present in the gut prior to infection are predictive of
\emph{C. difficile} colonization levels} Previous work in our group has
demonstrated that the microbes present in the mouse microbiome prior to
\emph{C. difficile} challenge can predict future colonization levels
(cite Alyx). To determine if the human-associated microbiome in the
mouse could similarly predict \emph{C. difficile} colonization we
employed a Random Forest machine learning algorithm similar to the one
used in previous studies. Relative abundance of OTUs with greater than
1\% abundance at day 0 were used as input to the model. We found that
the model explained X\% of the variance in colonization levels of
\emph{C. difficile} (Fig 3A). By refining the model to include only the
top X predictive OTUs, the model explained X\% of the variance. Partial
dependency plots of the top 6 OTUs predictive of \emph{C. difficile}
colonization reveal that colonization is positively associated with
\emph{Bacteriodetes}, \emph{Parabacteriodes} and
\emph{Porphyromonadaceae} (really?) (Fig 3B). Colonization was
negatively associated with the presence of \emph{Lachnospiraceae} (Fig
3B) though none of the human-associated mouse communities were
completely resistant to colonization by \emph{C. difficile} (Fig 2A,
3B). These results confirm that as in mouse microbiome studies, the
human-associated gut community prior to challenge is predictive of
\emph{C. difficile} colonization levels.

@@

Finally, our pilot experiments testing the efficiency of infection with
different \emph{C. difficile} strains

Discuss different strain results

• Future work/conclusion section/sentence

@@

\textbf{DNA isolation and sequencing} Bacterial DNA was isolated from
banked donor stool and each mouse stool pellet using the MO-BIO
PowerSoil DNA isolation kit. This DNA was used as template for

• Bacteria/plating • Sequencing • Data analysis, code availability •
Machine learning models

@@

Lab, sequencing core, Jhansi

>>>>

\emph{Clostridium difficile} causes diarrheal disease when it
successfully colonizes a disrupted gut microbial community. Current
mouse models to study \emph{C. difficile} infection (CDI) rely on
pre-treatment with antibiotics to alter the mouse microbiome prior to
inoculation. This model does not allow for analysis of human-associated
microbial community members that modulate \emph{C. difficile}
colonization and expansion. To study human-associated microbes in the
context of CDI, we inoculated germ-free C57BL/6 mice with one of 16
human fecal samples from diarrheal or healthy donors and challenged with
C. difficile 14 days later. Five unique donor-mice combinations resulted
in severe CDI while the remaining 11 only experienced mild disease. Both
healthy and diarrheal donors were susceptible to colonization and severe
symptoms of CDI. To determine if specific microbes were associated with
disease severity outcomes, we built a classification Random Forest
machine learning model based on relative abundance data of the
communities prior to infection. The model identified a number of
bacterial populations associated with the development of severe CDI,
including \emph{Bacilliales, Ruminococcaceae, Ruminococcus,
Staphylococcus, Streptococcus and Bacteriodetes}. Additionally, a
regression model accurately predicted colonization levels of \emph{C.
difficile} at one to ten days post-infection. This model explained 99\%
of the variance in the number of CFU isolated from mouse stool. Members
of \emph{Lachnospiraceae, Parabacteroides, Bacteroidales, Bacteroidetes,
Porphyromonadaceae} and unclassified \emph{Bacteria} families were
predictive of future \emph{C. difficile} colonization levels. Finally,
challenging these mice with different strains of \emph{C. difficile}
revealed that susceptible human-associated microbial communities were
prone to severe disease independent of strain type. Taken together these
results suggest that human-associated microbial communities can be
recapitulated in germ-free mice and used to characterize dynamics of
CDI. Because both healthy and diarrheal patients were susceptible to
severe CDI, machine-learning models are useful to identify bacterial
populations that allow colonization and contribute to the development of
\emph{C. difficile} associated disease in humans.

@@

The power of the gnotobiotic models to study CDI has been further
realized by first inoculating germ-free mice and piglets with human
stool microbes. In one study, germ-free piglets were acutely colonized
with human feces for one week and then treated with tigecycline. After
challenge with \emph{C. difficile} none of the antibiotic-treated
piglets succumbed to infection, while some of the untreated
human-colonized pigs did (11). In another study, germ-free mice
colonized with human feces were bred over several generations to create
a cohort of mice with identical human-derived microbiomes (14). These
mice were subsequently treated with a five-antibiotic cocktail to induce
dysbiosis and then were successfully colonized by \emph{C. difficile}
(14). While informative, these studies were limited in their use of only
one human donor as input inoculum.

Here we designed a laboratory system that allows for studying and
modeling microbial coumminity interactions with \emph{C. difficile}
infection in mice with human derived microbiomes. To test the impact of
individual human microbiomes on CDI, we colonized germ-free mice with 16
different human stool donors. We then characterized human-associated
microbiome response to \emph{C. difficile} challenge. Further, we built
a predictive model based on bacterial composition that classified
``at-risk'' microbiomes prior to infection with \emph{C. difficile}.
These findings show that human-associated microbiomes can be at risk for
CDI even in the absence of antibiotic-induced dysbiosis. Finally, our
results indicate that study of mice colonized with human feces can
replicate a range of clinical outcomes.

@@

\textbf{Germ-free mice inoculated with human feces as model for C.
difficile infection.} To generate mice with human-derived microbiomes,
we inoculated one cage each of gnotobiotic C57/BL6 mice with one of 16
different human fecal donors. Five donors were patients that had
diarrhea that was not attributable to \emph{C. difficile} infection
while 11 donors were healthy at time of donation. Stool from a patient
that was colonized with virulent \emph{C. difficile} was used as a
positive control. After inoculation with human stool, mice microbiomes
were allowed to equilibrate for 14 days. Prior to infection, stool
samples were taken from each mouse to establish baseline microbiome
composition. Then, 100 spores of \emph{C. difficile} strain 430
(isolated from the positive control patient) were used to orally infect
each mouse. Mice were monitored for weight loss and clinical signs of
disease. Fecal samples were taken to enumerate \emph{C. difficile} CFU
and for microbiome analysis each day up to 10 days post-infection (Fig
1A). To ensure that the donors we selected represented a diverse array
of human microbiomes, we sequenced the 16S rRNA genes from donor fecal
inocula. Ordination of the distances between donor communities showed
that the donors each had distinctly different communities, independent
of whether the sample came from a sick or healthy person (Fig 1B).
Likewise, the microbial communities of the mice on day 0 were
characterized by sequencing of fecal pellet DNA prior to infection
(Table S1). Ordination of day 0 communities shows that mice were similar
to each other within each cage and donor, but distinct from other donors
(Fig 1C). This result confirmed that human-associated microbes were able
to colonize gnotobiotic mice and provide various initial communities to
test \emph{C. difficile} dynamics.

\textbf{C. difficile infection in mice with human-derived microbiota
cause a range of outcomes.} \emph{C. difficile} colonization was
monitored by daily plating of stool pellets for \emph{C. difficile} CFU.
Nearly all of the mice were colonized to 10\textsuperscript{5} --
10\textsuperscript{7} CFU by one day post-infection and remained
colonized at that level until the end of the experiment (Fig 2A). As one
indicator of disease, mouse weights were taken each day post-infection
and weight-loss was monitored alongside clinical signs. When mice were
judged to be too ill to continue they were humanely euthanized. Overall,
disease phenotypes fell into two classes. Mice that became severely ill
and lost 20\% or more of their starting body weight within one to two
days post-infection were classified as ``severe'' whereas mice that were
colonized with \emph{C. difficile} but did not show signs of disease or
severe weight loss were considered to have ``mild'' disease (Fig 2A,
2B). Interestingly, \emph{C. difficile} was able to cause severe disease
in both mice that had been colonized with healthy stool and those
colonized with diarrheal stool, suggesting susceptibility to CDI is
dependent on the composition of the starting microbiome and not
associated with donor clinical metadata.

\textbf{Nick will be updating the next two sections} \textbf{Microbes
present in the gut prior to infection are predictive of \emph{C.
difficile} colonization levels} Previous work in our group has
demonstrated that the microbes present in the mouse microbiome prior to
\emph{C. difficile} challenge can predict future colonization levels
(cite Alyx). To determine if the human-associated microbiome in the
mouse could similarly predict \emph{C. difficile} colonization we
employed a Random Forest machine learning algorithm similar to the one
used in previous studies. Relative abundance of OTUs with greater than
1\% abundance at day 0 were used as input to the model. We found that
the model explained X\% of the variance in colonization levels of
\emph{C. difficile} (Fig 3A). By refining the model to include only the
top X predictive OTUs, the model explained X\% of the variance. Partial
dependency plots of the top 6 OTUs predictive of \emph{C. difficile}
colonization reveal that colonization is positively associated with
\emph{Bacteriodetes}, \emph{Parabacteriodes} and
\emph{Porphyromonadaceae} (really?) (Fig 3B). Colonization was
negatively associated with the presence of \emph{Lachnospiraceae} (Fig
3B) though none of the human-associated mouse communities were
completely resistant to colonization by \emph{C. difficile} (Fig 2A,
3B). These results confirm that as in mouse microbiome studies, the
human-associated gut community prior to challenge is predictive of
\emph{C. difficile} colonization levels.

@@

Finally, our pilot experiments testing the efficiency of infection with
different \emph{C. difficile} strains have allowed for testing community
versus pathogen factors of infection. Our results are consistent with
previous reports that the composition of the gut microbiome determines
if an individual is successfully infected or not (cite Alyx and others).
\emph{C. difficile} strains are known to vary in pathogenecity,
primarily via the amount of toxin that is produced during infection
(cite). In these pilot experiments, we challenged a susceptible
community with another pathogenic strain of \emph{C. difficile} to
confirm that the susceptibility lied in the community and not in the
specific challenge strain. In contrast, we exposed a
resistant/asymptomatic community to a relatively weak strain of \emph{C.
difficile} and found that the community could be colonized by this one
as well-- but not cause disease. This suggests that this colony is prone
to colonization by any \emph{C. difficile} strain. Further experiments
will test if infecting susceptible communities with weak strains allows
the mice to survive. It has been recently reported that carriage of a
non-toxigenic strain of \emph{C. difficile} can prevent additional
colonization by a pathogenic strain, suggesting some role of
niche-filling or immunity this provides (cite cite). It would be
interesting to see if we could make a susceptible community resistant to
infection and disease by a pathogenic \emph{C. difficle} strain in our
human-associated microbiome mice. Further work with human-associated
communities in mice and mathematical modeling will allow for uncovering
of complex human-associated disease processes and dynamics as they
relate to human disease.

@@

\textbf{DNA isolation and sequencing} Bacterial DNA was isolated from
banked donor stool and each mouse stool pellet using the MO-BIO
PowerSoil DNA isolation kit. This DNA was used as template for
amplification of the V4 region of the 16S rRNA gene and fragments were
sequenced on an Illumina MiSeq as previously described (cite Jim paper).

• Sequence curation/mothur, data analysis, code + availability of
data(SRA) and code (Github) - Nick

@@

We would like to thank members of the Schloss lab and Jhansi Leslie for
analysis advice and the Microbial Systems Laboratory at the University
of Michigan for sequencing assistance.
